Untwisting the Boerdijk--Coxeter Helix

### 2017-04-24 doc/TetrahelixUntwist.tex
working on proofs

A \emph{triple helix} is three congruent helices that share an axis. We show that
optimal tetrahelices are in fact triple helices with the same radius, so that all
vertices are on a cylinder. In, stages, we demonstrate that optimal tetrahelices:
\begin{enumerate}
\item have the same pitch,
\item have parallel axes,
\item share the same axis,
\item have the same radius.
\end{enumerate}

@@

Since all three rails have the same rail edge length, no matter how we
move the rails in the $xy$ plane if we shorten the $xy$ distance between
vertices we shorten the total distance.
Our tool for thinking about this is to collapse the $z$ dimension to form
a two-dimensional figure of nodes and non-rail edges (see \cref{untouchingrailfig}.)
Consider the projection along the $z$ axis of all vertices and non-rail edges into the $xy$ plane, which will be
a figure of dots and connecting segments in the $xy$ plane. The convex
hull for any one helix projection will be a circle (if its pitch is
irrational) or a polygon if rational, or a point if the helix has
zero curvature. Each of these figures by definition lies outside the
circle of inradius $r_{in}$ in the $xy$ plane.

@@

Note that \cref{thm:cylinder} is not true when $\rho$ is an odd integer multiple of $\pi$.

Note: This is not a proof that they have the same radius, I will have to add that.

@@

 In principle in any 3 helices with the same axis of the same radius
 having any relative
 displacement along the $z$ axis there are 9 distinct edge classes.
 If when projecting all vertices on the the $z$-axis, the interval
 defined by the $z$ axis value of its endpoints contains no other vertices,
 we call it a \emph{one-hop} edge, and if it does contain another vertex we
 call it a \emph{two-hop} edge.
 Then there are 
 3 rail edges, 3 one-hop lengths between each pair of 3 rails, and 3 two hop
 lengths between each pair of 3 rails, where the two-hop length is at least
 the one-hop length.
 However, we have already shown the rail
 lengths are equal in any optimal tetrahelix.

@@

By \cref{eventhirds} every optimal tetrahelix has vertices lying on helices expressible in the form:
\[
V_{optimal}(n,c) =
\left [
  \begin{tabular}{c}
   $ r \cos(n \alpha +  c 2 \pi /3)$\\
   $ r \sin(n \alpha +  c 2 \pi /3) $\\
   $ \frac{d(n +c / 3)}{3}   $
  \end{tabular}
  \right ],
\text{where:}
\begin{tabular}{c}
  $c \in \{0,1,2\}$
  \end{tabular}      
\]
where we have not yet investigated in the general case the relationships beteween $\alpha$, $r$, and $d$ in this formulation.
However, we understand that when $\alpha = 0$, the helices are degenerate, having curvature of $0$, and
we have the equitetrabeam.

\section{Parameterizing Tetrahelices via Rail Angle}

@@

\begin{theorem}
  \label{thm:cylinder}
  Any optimal tetrahelix with a rail angle less than $\pi$ has all vertices on a single cylinder.
\end{theorem}

\begin{proof}

Case 1: Suppose that $\rho$ is zero. Then for any given inradius, an equilateral
triangle is the minimax solution for all non-rail edges. Since there is only
one rail-edge length, this is the minimax solution for the entire set. Since
the vertices of an equilateral triangle lie on a circle, all points lie on
a cylinder.

Case 2: Suppose that $\rho$ is positive but less than $\pi$.
In this case each rail helix has
curavature and places points on both sides of any line through its axis.

We first show that the inradius is touched by one segment from each pair
of rails.

\begin{figure}[H]
     \centering
     \includegraphics[width=0.7\textwidth]{figures/UntouchingRail.png}
     \caption{Untouching Rail}
  \label{untouchingrailfig}
\end{figure}

\Cref{untouchingrailfig} depicts a projection of
only the point $P$, the point $P'$ which is an adjacent vertex on the same rail,
and the intermediary points $R$ and $Q$ on the other two rails.

Supoose there is a rail $P$ which does not have a segment touching the incircle
at all in the projection, as depicted in \cref{untouchingrailfig}.
Then a segment connecting the other two $Q$ and $R$ is a longest
rail, because any chord touching an incircle is longer than any which does not.
These two rails ($Q$ and $R$) can be moved closer to each other, and further away from $P$,
to form a better minimax solution
by shortening the longest rail, until one of the segments from $P$, the previously
untouching rail, touches the incircle. By induction, then all rails have
at least one segement touching each vertex on the rail that is tangent to the
incircle in an optimal tetrahelix.

Now suppose that you attempt to move the axis of one of the rails away from the center
for our circle. Because
$\rho > 0 $ and $\rho$ is not a multiple of $\pi$, there is some vertex on the two sides of any diameter
halving our circle. Moving the axis of a tetrahelix lengthens a longest line while ``pinching'' the
inradius on the other side, so it is not optimal. Since the axes of these helices are the
same and they have the same curvature and pitch, the points defined by the helices all lie
on a cylinder.
\end{proof}

>>>>

A \emph{triple helix} is three congruent helices that share an axis. We show that
optimal tetrahelices are in fact triple helices with the same radius, so that all
vertices are on a cylinder. In, stages, we demonstrate that optimal tetrahelices:
\begin{enumerate}
\item have the same pitch,
\item have parallel axes,
\item share the same axis,
\item have the same radius,
  \item have the same rail lengths,
  \item have axialy equidistant nodes, and therefore
  \item are in fact triple helices.
\end{enumerate}

@@

%% Since all three rails have the same rail edge length, no matter how we
%% move the rails in the $xy$ plane if we shorten the $xy$ distance between
%% vertices we shorten the total distance.
%% Our tool for thinking about this is to collapse the $z$ dimension to form
%% a two-dimensional figure of nodes and non-rail edges (see \cref{fig:wobble}.)
%% Consider the projection along the $z$ axis of all vertices and non-rail edges into the $xy$ plane, which will be
%% a figure of dots and connecting segments in the $xy$ plane. The convex
%% hull for any one helix projection will be a circle (if its pitch is
%% irrational) or a polygon if rational, or a point if the helix has
%% zero curvature. Each of these figures by definition lies outside the
%% circle of inradius $r_{in}$ in the $xy$ plane.

@@

Now we show that an optimal tetrahelix has the same radii for all three helices.
To do this exhibit a symmetric tetrahelix (not yet shown to be optimal) which
happens to be a tetrahelix, that has
the property that all rail edges are equal to all one-hop edges and all two-hop
edges are equal. Observe that although we have not get given the formula for the
radii of such a triple helix, we observe there are some values for $r$ and $\alpha$
for which this is true from this generic formula for a triple helix:
\[
\label{eq:triple}
V_{triple}(n,c) =
\left [
  \begin{tabular}{c}
   $ r \cos(n \alpha +  c 2 \pi /3)$\\
   $ r \sin(n \alpha +  c 2 \pi /3) $\\
   $ \frac{d(n +c / 3)}{3}   $
  \end{tabular}
  \right ],
\text{where:}
\begin{tabular}{c}
  $c \in \{0,1,2\}$
  \end{tabular}      
\]
In \cref{sec:parameterizing} we show how to find $r$ and $\alpha$.

In principle in any 3 helices with the same axis of the same radius
having any relative
displacement along the $z$ axis there are 9 distinct edge classes.
If when projecting all vertices on the the $z$-axis, the interval
defined by the $z$ axis value of its endpoints contains no other vertices,
we call it a \emph{one-hop} edge, and if it does contain another vertex we
call it a \emph{two-hop} edge.
Then there are 
3 rail edges, 3 one-hop lengths between each pair of 3 rails, and 3 two hop
lengths between each pair of 3 rails, where the two-hop length is at least
the one-hop length.
However, if we generate the three helices symmetrically with \cref{eq:triple},
many of these lengths will be the same.

Now consider a tetrahelix in which the radius of one of the helices is different.
By the connections made in a tetrahelix, any increase to a radius increases both
a one-hop and two-hop distance, and any decrease likewise decreases two.
Since there exists a tetrahelix which has only two distinct classes of edge lengths,
(the smaller being one-hop = rail, the larger being the two-hop distance), the helix
with a larger radius increases a longest edge without increasing a shortest edges.
Likewise, a helix with a smaller radius decreases a one-hop edge without decreasing
a two-hop edge.  Therefore, a tetrahelix with different radii is not as some
tetrahelix generated by \cref{eq:triple}, and so it not optimal. (Note however,
that we have not yet exhibited a truly optimal tetrahelix, nor proven that such
a tetrahelix occurs when the one-hop edge, length is equal to the rail edge length,
which we do in \cref{sec:parameterizing}.)

Because an optimal tetrhelix has equivalent radii for all three helices,
it has equivalent rail edge lengths.

%% \begin{theorem}
%%   \label{thm:cylinder}
%%   Any optimal tetrahelix with a rail angle less than $\pi$ has all vertices on a single cylinder.
%% \end{theorem}

%% \begin{proof}

%% Case 1: Suppose that $\rho$ is zero. Then for any given inradius, an equilateral
%% triangle is the minimax solution for all non-rail edges. Since there is only
%% one rail-edge length, this is the minimax solution for the entire set. Since
%% the vertices of an equilateral triangle lie on a circle, all points lie on
%% a cylinder.

%% Case 2: Suppose that $\rho$ is positive but less than $\pi$.
%% In this case each rail helix has
%% curavature and places points on both sides of any line through its axis.

%% We first show that the inradius is touched by one segment from each pair
%% of rails.

%% \begin{figure}[H]
%%      \centering
%%      \includegraphics[width=0.7\textwidth]{figures/UntouchingRail.png}
%%      \caption{Untouching Rail}
%%   \label{untouchingrailfig}
%% \end{figure}

%% \Cref{untouchingrailfig} depicts a projection of
%% only the point $P$, the point $P'$ which is an adjacent vertex on the same rail,
%% and the intermediary points $R$ and $Q$ on the other two rails.

%% Supoose there is a rail $P$ which does not have a segment touching the incircle
%% at all in the projection, as depicted in \cref{untouchingrailfig}.
%% Then a segment connecting the other two $Q$ and $R$ is a longest
%% rail, because any chord touching an incircle is longer than any which does not.
%% These two rails ($Q$ and $R$) can be moved closer to each other, and further away from $P$,
%% to form a better minimax solution
%% by shortening the longest rail, until one of the segments from $P$, the previously
%% untouching rail, touches the incircle. By induction, then all rails have
%% at least one segement touching each vertex on the rail that is tangent to the
%% incircle in an optimal tetrahelix.

%% Now suppose that you attempt to move the axis of one of the rails away from the center
%% for our circle. Because
%% $\rho > 0 $ and $\rho$ is not a multiple of $\pi$, there is some vertex on the two sides of any diameter
%% halving our circle. Moving the axis of a tetrahelix lengthens a longest line while ``pinching'' the
%% inradius on the other side, so it is not optimal. Since the axes of these helices are the
%% same and they have the same curvature and pitch, the points defined by the helices all lie
%% on a cylinder.
%% \end{proof}

%% Note: This is not a proof that they have the same radius, I will have to add that.

@@

 However, we have already shown the rail
 lengths are equal in any optimal tetrahelix.

@@

By \cref{eventhirds} we are justified renaming \cref{eq:triple} ``optimal'':
\[
V_{optimal}(n,c) =
\left [
  \begin{tabular}{c}
   $ r \cos(n \alpha +  c 2 \pi /3)$\\
   $ r \sin(n \alpha +  c 2 \pi /3) $\\
   $ \frac{d(n +c / 3)}{3}   $
  \end{tabular}
  \right ],
\text{where:}
\begin{tabular}{c}
  $c \in \{0,1,2\}$
  \end{tabular}      
\]
where we have not yet investigated in the general case the relationships beteween
$\alpha$, $r$, and $d$ in this formulation.
However, we understand that when $\alpha = 0$, the helices are degenerate, having curvature of $0$, and
we have the equitetrabeam.

\section{Parameterizing Tetrahelices via Rail Angle}
\label{sec:parameterizing}

@@

Proof Ideas:

Now that we have shown parallel and conincident, we can definie ``one-hop'' and ``two-hop''
clearly.

Can we show that one-hop < two-hop in an optimal tetrahelix?

This is true by definition, but we potentially have 6 such differences.
If all are less than the maximum rail edge, we are not optimal, as a lower rail edge
is better.
If all are > than the minimum rail edge, we are not opitimal, as we can raise that
rail edge.

So MAX(non-rail) >= MIN(rail-edge).
also: MIN(non-rail) <= MAX(rail).

Conjecture: Increasing Radius increases length of all attached nodes.
Lemma: If a line is drawn between a point P on one circule and a point Q on a nother circle,
and the circles are parallel in space,
increasing the radius of eitehr circle increases the length, no matter where the circles are,
and no matter where the points are on the circle.

Therefore if there is a single maximum radius and it has a longest edge coming to it,
the minimax is made better by decreasing that radius until equal with the next radius.

If a single radius is a shortest radius, the minimax is made better by
increasing it.

Can we proe this rigourously?

Does a single large radius necessarily have a longest edge?
A Rail edge would be longest if we allow any rotation, but we can severely
limit rail angle to \rho_{bc}.

New Idea:

Suppose Red is a largest radius.  Blue and Yellow must be symmetrically distributed
ccw and cw from Red.

Proof sketch: if Red is a largest radius, the two-hop edges to Yellow and Blue
are the longest two in the system.  Is one is longer than that other, rotating it
is more optimum.

Proof Idea:

If two-hop edges are always longest and one-hop always shortest, then they
should be equalized.

* * *

Proof idea.

Thm: An optimal tetrahelix has a longest edge which is a two-hop edge.

Proof:

Case 1: If the longest edge is a rail edge, construct a better minimax optimal
tetrahelix by increasing the radius of every helix, which does not change the
rail length, until a non-rail edge is equal to the rail edge.  This increases
the length of every non-rail edge, therefore it is a better helix.

Case 2: The longest edge is a non-rail edge.

WOLOG, this is an edge from Red to Blue. By assumption, at least one
of three things must be true:
1) The RED-BLUE 2-hop axial distance is longer than the other two 2-hop axial distances.
2) One rail (call it Red) has a larger radius.
3) The Red-Blue rails can be rotated closer together to decrease the
RED-BLUE distance. (If the angle between RED and BLUE is zero, then one
of the other must be true.)

In case 1), construct a better tetrahelix axially shifting the BLUE rails to decrease
the RED-BLUE 2-hop axial distance (and increase the 2-hop axial distance.) The effects
our classes by:
RR not affected
BB not affected
YY not affected
RB- increased
RB+ decreased
BY- decreased
BY+ increased
RY- not affected
RY+ not affected

So if the set of shortest edges does not include BY-, this tetrahelix is better.
If it does include BY-, we must show that RB+/BY- decreases as we move the rail.
We could show the actual derivative of this.  Informally, the angle is more direct,
so the change in RB+ should be greater. This may require comparison to the cylindrical
solution.  If we assume this is true for the cylindrical solution, then we have shown
a better solution exists. So case one is not optimal.

So we must show that if a shortest edge is in RY-, RY+, RB-, we have to
RB+ of these edges decreases. RY+ cannot be a shortest edge.
Idea: if RY-, RB- are shortest, increase Y, B radius.
So what does the derivative of RB+/RY- and RB+/RB- tell us? The further away it is,
the straighter the push, so it is likely that the RB+ edge changes faster than the RY- edge
decreases. If R has a bigger radius, can RY- be smaller than BY-? Seems like we will
have to use the algebraic formula to do this---but I have no proof yet that it is an triple helix.

Note, that if we could prove just this case, our equidistance proof would complete everything we need!
So I need to show that RB+/RY- decreases as we shrink the radius---in other words, the RB+ line
is more central than the RY- line.

In case 3), if we rotate blue towards red:

RR not affected
BB not affected
YY not affected
RB- decreases
RB+ decreases
BY- increases
BY+ increases
RY- not affected
RY+ not affected

Attempt to create Proof of Equal Radius:

Suppose that Red Radius is largest.

Case RB+ is longest:
We can understand the change of decreasing the Red Radius:

RR decreases
BB not affected
YY not affected
RB- decreases
RB+ decreases
BY- not affected
BY+ not affected
RY- decreases
RY+ decreases

Sub Case split:
Case RB- is smallest:
Increase B radius.
Case RY- is smallest:
Increase Y radius.
Case RR is smallest:
This is not possible if Red is has the largest Readius!
Case: RB- = RY-  are smallest.  In increase both B and Y.

Done case RB+ is longest.

Case YB+ is longest:

This can occur only if YB is not even 1/3rds.

Case:
YB- is the worst. Then we need to move the bar.

Case:
Rail BB is smallest.  Then  we are making better.

Case:
RB- is smallest (need to move blue.)
RY- is smallest (need to move yellow.)

(I think this is a combined proof.)

Case RB+ amd YB+ are both longest.

RR decreases
BB not affected
YY not affected
RB- decreases
RB+ decreases
BY- not affected
BY+ not affected
RY- decreases
RY+ decreases

Decrease R and simultaneously rotate Y towards B
to decrease YB+.

Done with case YB+ is longest.

Case RR is longest:

Decrease radius improves.

Now suppose the Red and Blue Radiii are equual and larger than yellow.

Case split RB+ is longest.

Increase Y does:
RR not affected
BB not affected
YY increases
RB- not affected
RB+ not affected
BY- increases
BY+ increases
RY- increases
RY+ increases

Sub case split: Shortest not RR,BB, or RB- or RB+, then we improve.
If it is a rail, decrease all radii until this is not true.

RB-, then we need to move Red relative to Blue. If we are already
at even thirds and proper rotation is this possible? This might require
going down the algebra level to prove.

Case split YB+ or YR+ is longest.

New Proof Idea:

Prove optimality of equal one-hop edges from simple diagram.
From that, show optimality of equidistance and radius.

New Proof Idea:

Proof that all radii are the same.

Lemma: There is a solution with all one-hop edges equal to all rail edges
and all two-hop edges longer.

Assume that the solution with all one-hop and rail distances the same
is optimal.  Assume that all two-hop edges are the same.

Any change to a radii either increases both or decreases both for
one set, thereby harming the optimum.  Y cannot be rotated
toward both Red and Blue. Y cannot be shifted toward both Red and Blue.
(We need a lemma here that Red and Blue are on the opposite side or the
line through the circle of Red.)
Therefore any different radius has a longer two-hop edge, and is not optimal.

If we have all radii the same, then our existing proof works for
equidistance.
